\documentclass[12pt]{article}
\usepackage[top=0.9in, bottom=0.5in, left=0.9in, right=1.1in]{geometry}

\usepackage{graphicx,color,enumitem}
\usepackage{amsmath,amsthm,amsbsy}
\usepackage{palatino}

%% Setup aproblem environment, 
%% aproblem items
%% subproblems environment
%% subproblem items
\makeatletter
\newcounter{probcount}
\newcounter{subprobcount}
\newlength\probsep
\newlength\pshrinking
\newif\iffirstprob
\newenvironment{aproblems}%
  {\ifhmode\unskip\par\fi\setcounter{probcount}{0}\probsep\parskip
  \sbox\@tempboxa{\textbf{9.}}\pshrinking\wd\@tempboxa\advance\pshrinking\labelsep
  \let\hproblem\aproblem
  \advance\linewidth -\pshrinking
  \advance\@totalleftmargin\pshrinking
  \advance\leftskip\pshrinking}%
  {\ifhmode\unskip \par\fi\advance\leftskip-\pshrinking}%

\newcommand{\aproblem}{%
  \setcounter{subprobcount}{0}%
  \stepcounter{probcount}%
  \def\@currentlabel{\arabic{probcount}}%
  \ifhmode
    \unskip \par
  \fi
%  \addpenalty{-4000}%
  \iffirstprob\else\addvspace\probsep\fi
  \firstprobfalse
  \hskip -\labelwidth\hskip -\labelsep 
  \hbox to\labelwidth{\hss\textbf{\arabic{probcount}.}}\hskip\labelsep
}%

\newcommand{\subprob}{\item\def\@currentlabel{\arabic{probcount}\alph{\thelistlabel}}}
\newcommand{\skipproblem}{\stepcounter{probcount}}


%% The following commands put defined left and right headers on the top, and a page number
%% on the bottom of all pages beyond page 1
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyfoot[C]{\ifnum \value{page} > 1\relax\thepage\fi}
\fancyhead[L]{\ifx\@doclabel\@empty\else\@doclabel\fi}
\fancyhead[R]{\ifx\@docdate\@empty\else\@docdate\fi}
\headheight 15pt
\def\doclabel#1{\gdef\@doclabel{#1}}
\def\docdate#1{\gdef\@docdate{#1}}
\makeatother

%% General formatting parameters
\parindent 0pt
\parskip 6pt plus 1pt


\doclabel{Math F251: Sections 3.9 and 3.8 Worksheet}
\docdate{Friday 8 March 2019}


\begin{document}
\renewcommand{\d}{\displaystyle}

\begin{aproblems}
% 3.9 #13, 14
\noindent For each problem,
\renewcommand{\labelenumi}{(\alph{enumi})}

\vspace{-5mm}
\begin{enumerate}
\setlength\itemsep{0em}
\item Draw a sketch of the situation.
\item Name (as variables) the quantities which are changing in time.
\item Write an equation relating the (variable and constant) quantities.
\item Finish solving the problem.
\end{enumerate}

\aproblem A plane flying horizontally at an altitude of 1 mile and a speed of 500 miles per hour passes directly over a radar station.  Find the rate at which the distance from the plane to the station is increasing when it is 2 miles away from the station.

\vfill
\aproblem If a snowball melts so that its surface area decreases at a rate of 1 cm$^2/$min, find the rate at which the diameter decreases when the radius is 5 cm.

\vfill

\clearpage\newpage
% 3.8 #19
\aproblem  The rate of change of atmospheric pressure $P$ with respect to altitude $h$ is proportional to $P$.  (This assumes the temperature is constant.)

\renewcommand{\labelenumi}{(\alph{enumi})}
\begin{enumerate}
\item  Write a differential equation corresponding to the first sentence above; use $k$ for the constant of proportionality.  Then write a formula for $P(h)$ in terms of $P(0)$, $k$, and $h$.

\vfill
\item  At a temperature of $15\,^\circ C$, the pressure is $101.3$ kPa at sea level and the pressure is $87.14$ kPa at $h=1000$ m.  From these facts, determine $P(0)$ and $k$.

\vfill
\item  What is the pressure at the top of Denali, at an altitude of $6187$ m?  (\emph{This problem in the book, \#19 in \S 3.8, has an error.  It calls it ``Mount McKinley.''})

\vfill
\item  At what altitude is the pressure $1/3$ of what it is at sea level?

\vfill
\end{enumerate}

\end{aproblems}

\end{document}
