\documentclass[11pt]{amsart}
\usepackage[margin=1in, head=1in]{geometry}

\usepackage{fancyvrb,xspace}
\usepackage[pdftex,colorlinks=true,urlcolor=blue]{hyperref}

\parindent=0in
\parskip=0.5\baselineskip

%\fvset{fontsize=\small, numbers=left} 
\DefineShortVerb{\|}

\newcommand{\mfile}[1]{
\begin{quote}
\VerbatimInput[frame=single,framesep=3mm,label=\fbox{\normalsize \textsl{\,#1\,}},fontfamily=courier,fontsize=\scriptsize]{#1}
\end{quote}
}

\newcommand{\Matlab}{\textsc{Matlab}\xspace}
\newcommand{\Octave}{\textsc{Octave}\xspace}
\newcommand{\python}{\textsc{Python}\xspace}
\newcommand{\ipython}{\textsc{ipython}\xspace}
\newcommand{\pylab}{\textsc{pylab}\xspace}
\newcommand{\scipy}{\textsc{scipy}\xspace}
\newcommand{\matplotlib}{\textsc{matplotlib}\xspace}

\begin{document}

\title{Comparison of \textsc{Matlab/Octave} versus \textsc{pylab} versus {julia}}

\author{Ed Bueler}

\date{\today.}

\maketitle
\normalsize
\thispagestyle{empty}

\newcommand{\hrf}[2]{\href{#1}{\texttt{#2}}}

\Matlab (\hrf{http://www.mathworks.com/}{www.mathworks.com}) was designed by Cleve Moler around 1980 for teaching numerical linear algebra without needing FORTRAN.  It has become a powerful programming language and engineering tool.  More than half of Math 615 students are already familiar with it.  It is available in most labs and graduate student offices at UAF.

But I like free, open source software.  There are free alternatives to \Matlab, and they'll work well for this course.  First, \Octave is a \Matlab clone.  The examples below work in an identical way in \Matlab and in \Octave.  I will mostly use \Octave myself for teaching, but I'll test examples in both \Octave and \Matlab.  To download \Octave, go to 
\hrf{http://www.gnu.org/software/octave/}{www.gnu.org/software/octave}.

Second, the \scipy (\hrf{http://www.scipy.org/}{www.scipy.org}) and \matplotlib (\hrf{http://matplotlib.org/}{matplotlib.org}) libraries give the general-purpose interpreted language \python (\hrf{http://python.org/}{python.org}) all of \Matlab functionality plus quite a bit more.  This combination is called \pylab.  Using it with the \textsc{ipython} interactive shell (\hrf{http://ipython.org/}{ipython.org}) gives the most \Matlab-like experience.  However, the examples below hint at the computer language differences and the different modes of thought between \Matlab/\Octave and \python.  Students who already use \python will like this option.

On the next page are two algorithms each in \Matlab/\Octave form (left column) and \pylab form (right column).  To download these examples, follow links at the class page

\centerline{\hrf{http://bueler.github.io/M615S14/}{bueler.github.io/M615S14/}.}

Here are some brief ``how-to'' comments for the \Matlab/\Octave examples: \texttt{expint.m} is a \emph{script}.  A script is run by starting \Matlab/\Octave, usually in the directory containing the script you want to run.  Then type the name of the script at the prompt, without the ``.m'':

\verb|>> expint|

\noindent The second algorithm \verb|bis.m| is a \emph{function} which needs inputs.  At the prompt enter
\begin{Verbatim}
>> f = @(x) cos(x) - x
>> bis(0,1,f)
\end{Verbatim}
for example.  Doing \verb|help expint| or \verb|help bis| shows the block of comments as documentation.

For the \python versions:  Type \verb|run expint.py| at the \ipython prompt or \verb|python expint.py| at a shell prompt.  For the function \verb|bis.py|, run \python or \ipython and do: \verb|from bis import bis|.  In \ipython you can type \verb|bis?| to get documentation for that function, and run the example as shown in the docstring.

FIXME: COMMENTS ON JULIA

\vfill


\newpage
\vfill

\begin{tabular}{ll}
\hspace{-0.45in}\parbox{3.6in}{\mfile{expint.m}} & \hspace{-0.4in}\parbox{3.8in}{\mfile{expint.py}}
\end{tabular}

\begin{tabular}{ll}
\hspace{-0.45in}\parbox{3.6in}{\mfile{bis.m}} & \hspace{-0.4in}\parbox{3.8in}{\mfile{bis.py}}
\end{tabular}

\vfill

\newpage
\vfill

\begin{tabular}{ll}
\hspace{-0.45in}\parbox{3.6in}{\mfile{expint.jl}} & \hspace{-0.4in}\parbox{3.8in}{\mfile{bis.jl}}
\end{tabular}

\vfill

\end{document}

